---

title: "Visualizing LAPOP Data: 2023 AmericasBarometer Report"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing LAPOP Data: 2023 AmericasBarometer Report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_chunk$set(fig.width = 12, fig.height = 6)

```

## Introduction

This vignette demonstrates how to use the `lapop` package to reproduce figures from the 2023 *Pulse of Democracy* report.

[Read the full report here](https://www.vanderbilt.edu/lapop/ab2023/AB2023-Pulse-of-Democracy-final-20240604.pdf)

## Load Packages

```{r, packages}
library(lapop)
```

```{r, load data}
# AmericasBarometer 2007-2023 Country Merge (only Brazil)
data(cm23)

# AmericasBarometer 2023 Year Merge (only a subset)
data(ym23)
```

## Reverse Variables

```{r, reverse, eval=F}
cm23$aoj11r <- lpr_resc(cm23$aoj11, reverse = TRUE)
```

## Fonts

```{r, lapop fonts}
lapop_fonts()
```

## Figure 1.1: Time Series

```{r, time series}
fig1.1_data <- lpr_ts(cm23, outcome = "ing4", rec = c(5, 7), use_wave = TRUE)
lapop_ts(fig1.1_data,
         ymin = 50,
         ymax = 80,
         main_title = "Support for democracy decline a decade ago and remains comparatively low",
         subtitle = "% who support democracy")
```

## Figure 1.2: Cross-Country Comparison

```{r, cross country}
fig1.2_data <- lpr_cc(ym23, outcome = "ing4", rec = c(5, 7))
lapop_cc(fig1.2_data,
         main_title = "In many countries, only about one in two adults support democracy",
         subtitle = "% who support democracy")
```

## Figure 2.1: Multi-Line Over Time

```{r, mline}
fig2.1_data <- lpr_mline(cm23,
                         outcome = c("b13", "b21", "b31"),
                         rec = c(5, 7), rec2 = c(5, 7), rec3 = c(5, 7))
lapop_mline(fig2.1_data,
            main_title = "Trust in executives has declined to a level similar to other political institutions",
            subtitle = "% who trust...")
```

## Figure 2.3: CC Multiple Variables

```{r, ccm}
fig2.3_data <- lpr_ccm(ym23,
                       outcome_vars = c("b12", "b18"),
                       rec1 = c(5, 7),
                       rec2 = c(5, 7))
fig2.3_data$var <- ifelse(fig2.3_data$var == "b12", "Armed Forces", "National Police")
lapop_ccm(fig2.3_data,
          main_title = "The public typically reports more trust in the armed forces than the police, \nbut levels vary",
          subtitle = "% who trust...")
```

## Figure Mover Example: Satisfaction with Democracy

```{r, mover}
ym23$variables <- ym23$variables %>%
  mutate(edrer = case_when(
    edre <= 2 ~ "None/Primary",
    edre %in% c(3, 4) ~ "Secondary",
    edre > 4 ~ "Superior",
    TRUE ~ NA_character_
  ),
  q1tc_r = ifelse(q1tc_r == 3, NA, q1tc_r)) %>%
  mutate(q1tc_r = case_when(
    q1tc_r == 1 ~ "Men",
    q1tc_r == 2 ~ "Women",
     TRUE ~ NA_character_
  ))

attributes(ym23$variables$wealth)$label <- "Wealth"

fig_mover <- lpr_mover(ym23,
                       outcome = "pn4",
                       grouping_vars = c("q1tc_r", "edrer", "wealth"),
                       rec = c(1, 2))

fig_mover$varlabel <- ifelse(fig_mover$varlabel == "edrer", "Education", fig_mover$varlabel)

lapop_mover(fig_mover,
            main_title = "Satisfaction with democracy is significantly lower among women,
            those with higher \neducational attainment, and the middle class",
            subtitle = "% who are satisfied with democracy")
```

## FIgure Histogram: Citizens’ Views

```{r, histogram}
spot32_data <- lpr_hist(ym23, outcome = "vb21n")
spot32_data$cat <- c("Vote", "Run for\noffice", "Protest", "Participate\nin local orgs.",
                     "Other", "Change is\nimpossible")

lapop_hist(spot32_data,
           main_title = "On average in the LAC region, one in three say voting is 
           the best way to influence change",
           subtitle = "How do you believe you can best influence change?")
```

## Figure 3.5: Dumbell Plot

```{r, dumbell}
fig3.5_data <- lpr_dumb(cm23,
                        outcome = "q14f",
                        rec = c(1, 1),
                        over = c(2018, 2023),
                        xvar = "pais_lab",
                        ttest = TRUE)
fig3.5_data <- fig3.5_data[1:6, ]

lapop_dumb(fig3.5_data,
           main_title = "Among those with emigration intentions, the percentage who say 
           they are very likely\nto emigrate increased in Nicaragua and Guatemala",
           subtitle = "% who say it is very likely they will emigrate")
```

## Figure 3.8: Stack Plot by Country

```{r, stack}
fig3.8_data <- lpr_stack(ym23,
                         outcome = "q14f",
                         xvar = "pais",
                         order = "hi-lo")

lapop_stack(fig3.8_data,
            xvar = "xvar_label",
            source = ", AmericasBarometer 2023",
            main_title = "Nicaragua has the highest percentage of individuals with migration intentions, \nwhile Haiti has the lowest")
```

## End

Use this vignette as a template for your own reports and exploratory visualizations using LAPOP data.
